`forestecology` package for modeling interspecies competition between trees

### 2021-02-02 etc/paper/forestecology.Rmd
incorporated simon's suggestions

```{r load-packages}
library(forestecology)

# Load tidyverse packages:
library(tidyverse)
library(lubridate)

# Load spatial packages:
library(blockCV)
library(sf)
library(sfheaders)

# Load other packages:
library(snakecase)
library(yardstick)
```

@@

```{r bw-load-data}
data(census_2008_bw, census_2014_bw, species_bw)

# Append additional species data
census_2008_bw <- census_2008_bw %>%
  left_join(species_bw, by = "sp") %>%
  select(-c(genus, species, latin))
```

@@

```{r scbi-load-data}
scbi_2013 <- read_csv("scbi.stem2.csv") %>%
  select(treeID, stemID, sp, ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = ExactDate,
    dbh = as.numeric(dbh),
    date = mdy(date)
  ) %>%
  filter(gx < 300, between(gy, 300, 600)) %>% 
  mutate(dbh = dbh / 10)

scbi_2018 <- read_csv("scbi.stem3.csv") %>%
  select(treeID, stemID, sp, ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = ExactDate,
    dbh = as.numeric(dbh),
    date = mdy(date)
  ) %>%
  filter(gx < 300, between(gy, 300, 600)) %>% 
  mutate(dbh = dbh / 10)
```

@@

```{r bw-compute-growth}
census_2014_bw <- census_2014_bw %>%
  filter(!str_detect(codes, "R"))

bw_growth_df <-
  compute_growth(census_2008_bw, census_2014_bw, id = "treeID") %>%
  # Convert all variables denoting species to factors
  mutate(
    sp = sp %>% to_any_case() %>% as.factor(),
    species = sp,
    family = as.factor(family),
    trait_group = as.factor(trait_group)
  ) %>%
  # Drop unnecessary variables
  select(-stemID)
```

@@

```{r scbi-compute-growth}
scbi_growth_df <-
  compute_growth(scbi_2013, scbi_2018, "stemID") %>%
  # Convert all variables denoting species to factors
  mutate(sp = as.factor(sp))
```

@@

```{r growth-histogram, out.width="100%", fig.cap="Distribution of average annual growth in DBH for both sites.", echo = FALSE}
# Both Big Woods & SCBI
bind_rows(
  bw_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "Big Woods"),
  scbi_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "SCBI")
) %>% 
  ggplot(aes(x = growth, y = ..density.., fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  labs(x = "Average annual growth in dbh (cm per yr)") +
  coord_cartesian(xlim = c(-0.5, 1))
```

@@

bw_growth_df <- bw_growth_df %>%
  add_buffer_variable(direction = "in", size = comp_dist, 
                      region = study_region_bw)

ggplot() +
  geom_sf(data = bw_growth_df %>% sample_frac(0.2), 
          aes(col = buffer), size = 0.5)
```

@@

```{r bw-define-cv-folds, out.width="100%", fig.cap="blockCV function output."}
set.seed(76)
bw_spatialBlock <- spatialBlock(
  speciesData = bw_growth_df, theRange = cv_fold_size, k = 28, xOffset = 0.5, 
  yOffset = 0, verbose = FALSE, showBlocks = FALSE
)
```

@@

```{r}
bw_growth_df <- bw_growth_df %>%
  mutate(foldID = bw_spatialBlock$foldID)
```

```{r bw-define-cv-folds-with-trees, out.width="100%", fig.cap="Inspect blocks closely."}
# Visualize grid. Why does fold 19 repeat?
ggplot() +
  geom_sf(data = bw_spatialBlock$blocks %>% st_as_sf()) + 
  geom_sf(data = bw_growth_df %>% sample_frac(0.2), 
          aes(col = factor(foldID)), size = 0.1, show.legend = FALSE) + 
  geom_sf_text(data = bw_spatialBlock$blocks %>% st_as_sf(), 
               aes(label = folds))
```

@@

```{r}
bw_growth_df <- bw_growth_df %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19)) %>%
  mutate(foldID = as.character(foldID))
```

Separately, we save the spatial cross-validation grid as an `sf_polygon` object `bw_cv_grid`

```{r}
bw_cv_grid <- bw_spatialBlock$blocks %>%
  st_as_sf()
```

@@

```{r scbi-define-buffer, out.width="50%", fig.cap="Buffer region for SCBI study region."}
scbi_study_region <- tibble(
  x = c(0, 300, 300, 0, 0),
  y = c(300, 300, 600, 600, 300)
) %>%
  sf_polygon()

scbi_growth_df <- scbi_growth_df %>%
  add_buffer_variable(direction = "in", size = comp_dist, 
                      region = scbi_study_region)

ggplot() +
  geom_sf(data = scbi_growth_df, aes(col = buffer), size = 0.5)
```

@@

```{r}
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600)) %>%
  sf_polygon() %>% 
  mutate(folds = 1)
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600)) %>%
  sf_polygon() %>% 
  mutate(folds = 2)
scbi_cv_grid <- bind_rows(fold1, fold2)
```

```{r scbi-define-cv-folds, out.width="50%", fig.cap="Buffer region for SCBI study region."}
scbi_spatialBlock <- spatialBlock(
  speciesData = scbi_growth_df, k = 2, verbose = FALSE, showBlocks = FALSE,
  # Note new arguments:
  selection = "systematic", blocks = scbi_cv_grid
)

# Add foldID to each tree
scbi_growth_df <- scbi_growth_df %>%
  mutate(foldID = scbi_spatialBlock$foldID)

ggplot() +
  geom_sf_text(data = scbi_growth_df, 
               aes(label = foldID, col = buffer)) +
  geom_sf(data = scbi_cv_grid, fill = "transparent")
```

@@

```{r bw-focal-vs-comp}
focal_vs_comp_bw <- bw_growth_df %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = bw_cv_grid, id = "treeID")
```

@@

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- scbi_growth_df %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = scbi_cv_grid, id = "stemID")
```

@@

```{r bw-model-fit}
posterior_param_bw <- focal_vs_comp_bw %>%
  fit_bayesian_model(prior_param = NULL)
```

@@

```{r bw-model-predict}
predictions <- focal_vs_comp_bw %>%
  predict_bayesian_model(posterior_param = posterior_param_bw) %>%
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))
```

@@

```{r bw-model-rmse}
predictions %>%
  yardstick::rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-model-shuffle-fit}
posterior_param_bw_shuffle <- focal_vs_comp_bw %>%
  fit_bayesian_model(prior_param = NULL, run_shuffle = TRUE)
```

```{r bw-model-shuffle-predict}
predictions_shuffle <- focal_vs_comp_bw %>%
  predict_bayesian_model(posterior_param = posterior_param_bw_shuffle) %>%
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))
```

```{r bw-model-shuffle-rmse}
predictions_shuffle %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-model-fit}
posterior_param_scbi <- focal_vs_comp_scbi %>%
  fit_bayesian_model(prior_param = NULL, run_shuffle = FALSE)
```

```{r scbi-model-predict}
scbi_growth_df_noCV <- focal_vs_comp_scbi %>%
  predict_bayesian_model(posterior_param = posterior_param_scbi) %>%
  right_join(scbi_growth_df, by = c("focal_ID" = "stemID"))
```

```{r scbi-model-rmse}
scbi_growth_df_noCV %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Predicted versus observed growth.", echo = FALSE}
ggplot(scbi_growth_df_noCV, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="80%", fig.cap="Spatial distribution of residuals for model applied to SCBI data.", echo = FALSE}
scbi_growth_df_noCV %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

```{r bw-spatial-cv}
cv_bw <- focal_vs_comp_bw %>%
  run_cv(comp_dist = comp_dist, cv_grid = bw_cv_grid) %>%
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))

cv_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-spatial-cv}
cv_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = scbi_cv_grid) %>%
  right_join(scbi_growth_df, by = c("focal_ID" = "treeID"))

cv_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-posterior-viz-0}
focal_vs_comp_bw <- bw_growth_df %>%
  mutate(sp = family) %>%
  create_focal_vs_comp(comp_dist = comp_dist, cv_grid_sf = bw_cv_grid, 
                       id = "treeID")

posterior_param_bw <- focal_vs_comp_bw %>%
  fit_bayesian_model(prior_param = NULL, run_shuffle = FALSE)
```

@@

```{r bw-posterior-viz-1}
posterior_plots_bw <- plot_bayesian_model_parameters(
  posterior_param = posterior_param_bw,
  sp_to_plot = c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
)
```

@@

```{r bw-posterior-viz-beta0, out.width="100%", fig.cap="Posterior distribution of beta0."}
posterior_plots_bw[["beta_0"]]
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh."}
posterior_plots_bw[["beta_dbh"]]
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for Big Woods."}
posterior_plots_bw[["lambda"]]
```

@@

```{r scbi-posterior-viz-1}
posterior_plots_scbi <- plot_bayesian_model_parameters(
  posterior_param = posterior_param_scbi,
  sp_to_plot = c("quru", "litu", "cagl", "cato")
)
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for SCBI."}
posterior_plots_scbi[["lambda"]]
```

>>>>

```{r load-packages}
library(tidyverse)
library(lubridate)
library(sf)
library(forestecology)
library(blockCV)
```

@@

```{r bw-load-data}
census_2008_bw <- census_2008_bw %>%
  left_join(species_bw, by = "sp") %>%
  select(-c(genus, species, latin))
```

@@

```{r scbi-load-data}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(treeID, stemID, sp, ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = ExactDate %>% mdy(),
    dbh = as.numeric(dbh)
  ) %>%
  filter(gx < 300, between(gy, 300, 600)) %>% 
  mutate(dbh = dbh / 10)

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(treeID, stemID, sp, ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = ExactDate %>% mdy(),
    dbh = as.numeric(dbh)
  ) %>%
  filter(gx < 300, between(gy, 300, 600)) %>% 
  mutate(dbh = dbh / 10)
```

@@

```{r bw-compute-growth}
growth_bw <-
  compute_growth(
    census_1 = census_2008_bw %>% 
      mutate(sp = to_any_case(sp) %>% factor()),
    census_2 = census_2014_bw %>% 
      filter(!str_detect(codes, "R")) %>% 
      mutate(sp = to_any_case(sp) %>% factor()),
    id = "treeID"
  )
```

@@

```{r scbi-compute-growth}
growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi %>% 
      mutate(sp = to_any_case(sp) %>% factor()),
    census_2 = census_2018_scbi %>% 
      filter(!str_detect(codes, "R")) %>% 
      mutate(sp = to_any_case(sp) %>% factor()),
    id = "stemID"
  )
```

@@

```{r growth-histogram, out.width="100%", fig.cap="Distribution of average annual growth in DBH for both sites.", echo = FALSE}
# Both Big Woods & SCBI
bind_rows(
  growth_bw %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "Big Woods"),
  growth_scbi %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "SCBI")
) %>% 
  ggplot(aes(x = growth, y = ..density.., fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  labs(x = "Average annual growth in dbh (cm per yr)") +
  coord_cartesian(xlim = c(-0.5, 1))
```

@@

growth_bw <- growth_bw %>%
  add_buffer_variable(direction = "in", size = comp_dist, region = study_region_bw)

ggplot() +
  geom_sf(data = growth_bw %>% sample_frac(0.2), aes(col = buffer), size = 0.5)
```

@@

```{r bw-define-cv-folds, out.width="100%", fig.cap="blockCV function output."}
set.seed(76)
bw_spatialBlock <- spatialBlock(
  speciesData = growth_bw, theRange = cv_fold_size, k = 28, xOffset = 0.5, 
  yOffset = 0, verbose = FALSE, showBlocks = FALSE
)
```

@@

```{r}
growth_bw <- growth_bw %>%
  mutate(foldID = bw_spatialBlock$foldID)
```

```{r bw-define-cv-folds-with-trees, out.width="100%", fig.cap="Inspect blocks closely."}
# Visualize grid. Why does fold 19 repeat?
ggplot() +
  geom_sf(data = bw_spatialBlock$blocks %>% st_as_sf()) + 
  geom_sf(data = growth_bw %>% sample_frac(0.2), 
          aes(col = factor(foldID)), size = 0.1, show.legend = FALSE) + 
  geom_sf_text(data = bw_spatialBlock$blocks %>% st_as_sf(), 
               aes(label = folds))
```

@@

```{r}
growth_bw <- growth_bw %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19)) %>%
  mutate(foldID = factor(foldID))
```

Separately, we save the spatial cross-validation grid as an `sf_polygon` object `blocks_bw`

```{r}
blocks_bw <- bw_spatialBlock$blocks %>%
  st_as_sf()
```

@@

```{r scbi-define-buffer, out.width="50%", fig.cap="Buffer region for SCBI study region."}
study_region_scbi <- tibble(
  x = c(0, 300, 300, 0, 0),
  y = c(300, 300, 600, 600, 300)
) %>%
  sf_polygon()

growth_scbi <- growth_scbi %>%
  add_buffer_variable(direction = "in", size = comp_dist, region = study_region_scbi)

ggplot() +
  geom_sf(data = growth_scbi, aes(col = buffer), size = 0.5)
```

@@

```{r}
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600))
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600))

blocks_scbi <- bind_rows(
  sf_polygon(fold1),
  sf_polygon(fold2)
) %>%
  mutate(folds = c(1, 2) %>% factor())
```

```{r scbi-define-cv-folds, out.width="50%", fig.cap="Buffer region for SCBI study region."}
SpatialBlock_scbi <- spatialBlock(
  speciesData = growth_scbi, k = 2, selection = "systematic", blocks = blocks_scbi,
  showBlocks = FALSE, verbose = FALSE
)

# Add foldID to each tree
growth_scbi <- growth_scbi %>%
  mutate(foldID = SpatialBlock_scbi$foldID %>% factor())

ggplot() +
  geom_sf_text(data = growth_scbi, aes(label = foldID, col = buffer)) +
  geom_sf(data = blocks_scbi, fill = "transparent")
```

@@

```{r bw-focal-vs-comp}
focal_vs_comp_bw <- growth_bw %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = blocks_bw, id = "treeID")
```

@@

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- growth_scbi %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = blocks_scbi, id = "stemID")
```

@@

```{r bw-model-fit}
comp_bayes_lm_bw <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL)
```

@@

```{r bw-model-predict}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat = predict(comp_bayes_lm_bw, focal_vs_comp_bw))
```

@@

```{r bw-model-rmse}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-model-shuffle-fit}
comp_bayes_lm_bw_shuffle <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)
```

```{r bw-model-shuffle-predict}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_bw_shuffle, focal_vs_comp_bw))
```

```{r bw-model-shuffle-rmse}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
```

@@

```{r scbi-model-fit}
comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL)
```

```{r scbi-model-predict}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat = predict(comp_bayes_lm_scbi, focal_vs_comp_scbi))
```

```{r scbi-model-rmse}
focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Predicted versus observed growth.", echo = FALSE}
ggplot(focal_vs_comp_scbi, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="80%", fig.cap="Spatial distribution of residuals for model applied to SCBI data.", echo = FALSE}
focal_vs_comp_scbi %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

```{r bw-spatial-cv}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_bw)

focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)

focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-posterior-viz-0}
focal_vs_comp_bw <- growth_bw %>%
  mutate(sp = family %>% factor()) %>%
  create_focal_vs_comp(comp_dist = comp_dist, cv_grid_sf = blocks_bw, id = "treeID")

comp_bayes_lm_bw <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL)
```

@@

```{r bw-posterior-viz-1}
sp_to_plot <- c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
```

@@

```{r bw-posterior-viz-beta0, out.width="100%", fig.cap="Posterior distribution of beta0."}
plot1 <- autoplot(comp_bayes_lm_bw, type = "intercepts")
plot1
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh."}
plot2 <- autoplot(comp_bayes_lm_bw, type = "dbh_slopes")
plot2
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for Big Woods."}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```

@@

```{r scbi-posterior-viz-1}
sp_to_plot <- c("quru", "litu", "cagl", "cato")
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for SCBI."}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```
